---
output: html_document
editor_options:
  chunk_output_type: console
---
# S.J.P. Eloff (10237161)
# Practical 3 - Exploratory Data Analysis
# Submission: 13 August 2018, 17:30

This practical is based on a  dataset containing the prices and other attributes of almost 54,000 diamonds.

```{r message=FALSE,cache=FALSE}
library("tidyverse")
library("ggplot2")
library("dplyr")
diamonds
```

## Variation

### Question 1 - Discrete variables

a. Visualise the variable 'cut' with a bar graph
b. The height of the bars displays how many observations occurred with each x value. Compute these values manually and assign the values to a variable 'cut_count'

### Answers 1
```{r, question_1}
## a. Amount of diamonds per cut type.
ggplot(diamonds,aes(cut)) +
  geom_bar()

## b. The manual method to calculate and display the data from above follows below.
summary_b <- diamonds %>% group_by(cut) %>% summarize(cut_count=n())
summary_b
```

### Question 2 - Continuous variables

a. Visualise the variable 'carat' with a boxplort
b. Visualise the variable 'carat' with a histogram using binwidth = 0.5
c. Compute a frequency table of the variable 'carat' (with the same binwidth) and assign the values to a variable 'carat_freq'
d. Filter diamonds with a size of less than three carats. Asign this filtered data to a variable 'smaller.
On the filtered variable 'smaller':
e. Overlay multiple histograms - one for each cut using geom_freqpoly() which produces line rather than bar histograms. Add a title to the plot.


### Answers 2
```{r, question_2}
#a. A boxplot representing the interesting relationship between carat and clarity.
ggplot(diamonds,aes(clarity,carat)) +
  geom_boxplot()

#b. A histogram depicting the amount of diamonds in each 0.5 binwidth carat class.
ggplot(diamonds,aes(carat)) +
  geom_histogram(binwidth=0.5)

#c. A frequency table of the carat variable.
binwidth = c(0,0.5,1,1.5,2,2.5,3,3.5,4,4.5,5)
summary_c <- cut(diamonds$carat,breaks=binwidth)
summary_c = factor(summary_c)
table(summary_c)

#d. Filtered data labeled as 'smaller'.
smaller <- filter(diamonds,carat<3)
smaller

#e. Histograms of carat frequencies for each cut.
ggplot(smaller,aes(carat,colour=cut)) +
  geom_freqpoly(binwidth=0.5)

```
 
## Covariation

### Question 3 -  A categorical and continuous variable
a. Visualise the variable 'smaller' for each cut using boxplots. Assign different colorus to each boxplot. Add a title to the plot.

b. Visualise the distribution of price by cut using geom_boxplot(). Assign different colors to each boxplot. Add a title to the plot.

### Answers 3
```{r, question_3}
#a. Carat statistics for each cut represented by boxplots.
ggplot(smaller,aes(cut,carat,colour=cut)) +
  geom_boxplot() +
  ggtitle("Carat statistics for each cut represented by boxplots")

#b. Price statistics for each cut represented by boxplots.
ggplot(smaller,aes(cut,price,colour=cut)) +
  geom_boxplot() +
  ggtitle("Price statistics for each cut represented by boxplots")

```


### Question 4 -  Two categorical variables
Consider the following plot of diamond cut vs color (both categorical variables):
```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))
```
The size of each circle in the plot displays how many observations occurred at each combination of values. Covariation will appear as a strong correlation between specific x values and specific y values.

a. Create an alternative visualisation for these two variables using geom_tile(). Change the colour palette to `spectral'

### Answers 4

```{r}
spectral_ds <- diamonds %>% count(cut,color)
ggplot(spectral_ds) +
  scale_colour_brewer(palette="Spectral") +
  geom_tile(aes(cut,color,fill=n))
```

### Question 5 - Two continuous variables
Visualise price vs carat using geom_boxplot().
- Assign diamond clarity to colour
- Reverse the direction of the scale so that the worst diamond clarity is assigned to dark blue and best diamond clarity is assignmed to white.
- Make the legend heading 'Diamond clarity'
- Make the plot background dark for better display

### Answers 5
```{r}
ggplot(diamonds) +
  scale_color_brewer(palette="Blues",direction=-1) +
  geom_boxplot(aes(carat,price,colour=clarity)) +
  labs(colour="Diamond clarity") +
  ggtitle("Price plotted against carat (by clarity, shown in colour)") +
  theme_dark()
```

