---
title: "Chapter 3 and 4 Practical A"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

# Chapter 3: 

Consider the Pareto$(\alpha, \lambda)$  distribution for the data set in the associated excel file provided.

a) Import the data into R.

```{r}
library(xlsx)
mydata <- read.xlsx("ParetoDataSet.xlsx",sheetIndex=1,header=FALSE)
```

b) Derive theoretical formulas for the estimates for $\alpha$  and $\lambda$, using the method of moments. Type your answer here using Latex.

c) Determine numerical values for the estimate in (b) for the imported data. 

```{r}
sumx <- mydata %>% summarize(sum(mydata))
sumx2 <- mydata %>% summarize(sum(mydata^2))
n <- mydata %>% summarize(n())
xbar <- sumx/n
xbar2 <- xbar*xbar;
alpha_A <- 2*(sumx2/n-xbar2)/(sumx2/n-2*xbar2)
alpha_A
lamda_A <- (alpha_A-1)*xbar
lamda_A
```

# Chapter 4:

Consider an excess-of-loss reinsurance arrangment with level $M=1500$. The underlying claims $X$ are assumed to be N($1000, 50^2$).

Simulate 100 values of the inurer's losses $Y$, as well as the corresponding 100 values of the reinsurer's losses $Z$. 

```{r}


```