---
output: html_document
editor_options:
  chunk_output_type: console
---
# Practical 5 - Programming
# Submission: 3 September 2018, 17:30
# SJP Eloff (Student nr. 10237161)

## Question 1: Check if a number is positive, negative or zero
A number entered by a user is checked whether its a positive number or a negative number or zero. The output of the code is a print statement that prints the type of number (positive, negative or zero).

In order to do this, you need to understand if statements, variables and constants and operators.

```{r, question_1}
num <- -5

if(num>0){
  print("positive")
}else{
if(num==0){
  print("zero")
}else{
  print("negative")
}
}

```
## Question 2: Check wheter an integer is a prime number
A positive integer greater than 1 which has no other factors except 1 and the number itself is called a prime number.
Step 1 is to check if the number is greater than 1
Step 2 is to check if the number is exactly divisible by any number from 2 to number - 1. If we find a factor in that range, the number is not prime. Else the number is prime.

The output of the code is a print statement that prints whether the number is a prime number or not.

For this question you will need the if...else statement, for loop, break and next statement.

```{r, question_2}
num <- 53 

output=0
# check if num is greater than 1
if(num>1){
# check num's factors
  output=1
  for(i in 2:(num-1)){
    if ((num%%i)==0) {
    output=0
    break
    }
  }
}

# output the result
if(num==2) output=1

if(output==1){
  print(paste(num,"is prime"))
}else{
  print(paste(num,"is not prime"))
}
```

## Question 3: Find the factorial of a number
The factorial of a number is the product of all the integers from 1 to that number.
Step 1: Check if the number is negative, positive or zero.
Step 2: If the number is negative, print: "Sorry, factorial does not exist for negative numbers"
Step 3: If the number is zero, print: "The factorial of 0 is 1"
Step 4: For all other numbers, print the factorial of the number.

For this question you will use the if...else Statement and the for Loop.

```{r, question_3}
num <- 12
factorial <- 1

# check if negative, positive or zero
if(num<0){
  print("Sorry, factorial does not exist for negative numbers")
}else if(num==0){
  print("The factorial of 0 is 1")
}else{
  for(i in 1:num){
    factorial=factorial*i
  }
  print(paste("The factorial of",num,"is",factorial))
}
```
## Question 4: Check for a leap year
A leap year is exactly divisible by 4 except for century years (years ending with 00). The century year is a leap year only if it is perfectly divisible by 400.
The output is a print statement print whether the year (user input) is a leap year or not.
For this question you will use the if...else Statement and operators

```{r, question_4}
year <- 3000 

# nested ifs for each restriction
if((year%%4)==0){
  if((year%%100)==0){
    if((year%%400)==0){
      print(paste(year,"is a leap year"))
    }else{
      print(paste(year,"is not a leap year"))
    }
  }else{
    print(paste(year,"is a leap year"))
  }
}else{
  print(paste(year,"is not a leap year"))
}

```

## Question 5: Find the greatest common divisor
The greatest common divisor (GCD) of two numbers is the largest positive integer that perfectly divides the two given numbers.
For example, the H.C.F of 12 and 14 is 2.
The output of this question is a print statement that prints the GCD of the two input numbers.
For this question you will use: functions, for Loop, if...else Statement, while loop

```{r, question_5}
num1 <- 18 
num2 <- 6

f_gcd <- function(x,y){
  if(x>y){
    small=y
  }else{
    small=x
  }
  for(i in 1:small){
    if((x%%i==0)&&(y%%i==0)){
      ans=i
    }
  }
  return(ans)
}

# output, calling the function
print(paste("The GCD of",num1,"and",num2,"is",f_gcd(num1,num2)))
```


